Child Signals of Need in Utila, Honduras

### 2024-09-02 Paper.qmd
Added more to the discussion.

## Positive assoication between child allopaenting effort and number of older children

@@

The effect of the number of younger children in the household on the frequency of a child's alloparenting effort. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-alloparenting_dep_var for the corresponding regression table.
:::

@@

There were significant negative relationships between summed signal cost (`r stats$mBFc$BodyFat$str`) crying frequency (`r stats$mBFcry$BodyFat$str`), and tantrum frequency (`r stats$mBFtantrum$BodyFat$str`) and a scaled composite of subscapular and tricep skinfold thicknesses when controlling for the sex of the child, the age of the child, the frequency of conflict with one's primary caretaker, household demographics,caretaker perceptions of child immune function, and an alloparenting effort by sex interaction. See @tbl-bodyfat_models for regression statistics and @fig-effect_bodyfat for effects plots. Subscapular and tricep skinfold thicknesses were also significant predictors of summed signal cost and tantrum frequency when included in the same models as the sole measure of body fat. See xxxxSI TABLESxxxx for regression tables.

@@

@@

There was no significant association between the frequency of child sadness and age. Part of this likely results from the fact that, unlike crying, sadness is thought to have important non-signaling functions including drawing one’s attention to a problem, motivating deeper analysis of one’s situation, and causing withdrawal from situations with low fitness prospects (cite), from which others can still obtain cue-based information about one's emotional state, well-being, and desires. By asking about sadness as a whole, our measure was unable to disentangle information caretakers gleaned from cues from information that came from signals. For this reason, a change in non-crying signals that arise from sadness (e.g., facial sadness and some, but all, forms of withdrawal) may have been harder to detect with our sample size even if young children expressed facial sadness more often. It is also possible that there is no effect of age on sadness frequency among the age ranges examined it. add in depression over the lifecourse

@@

Consistent with signaling models which emphasize the role of conflicts of interest in favoring the use of costlier signals (cite this lab and others), there were positive associations between caretakers' perceptions of conflict frequency and 1) the summed costliness of children's signaling, 2) the frequency of child signals of need, and 3) the frequency of children's most common need signaling strategy. Although conflicts of interest do not always lead to behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest, making it a useful proxy for the underlying conflicts of interest that may increase the benefits of deception and the costs of being deceived\[\[5\]\]. If this is the case, the association between signaling and conflict may represent children's attempts at credibly signaling their value of additional investment or the costliness of punishment to parents who would have good reason to doubt the trustworthiness of simple verbal request (or the trustworthiness of facial sadness and crying for those resorting to temper tantrums).

When viewed in isolation, the frequencies of sadness, crying, and throwing temper tantrums were also positively associated with conflict frequency between child and caretaker. However, the frequency of running away was not significantly associated with conflicts with parents. A retrospective study of parent-offspring conflict among Chon Chuuk immigrants in the northwest United States found that running away was the most common response to long-standing conflicts with parents and often resulted in beneficial outcomes for the child [@syme2023a]. The threat running away poses to child wellbeing and, therefore, its ability to credibly signal information about one’s current situation is likely highly dependent on one’s socioecology and age. Among the Chon Chuuk, running away is often seen as a suicide threat [@fischer1950; @syme2023a], likely due to xxxxxxxxx. Similar risks may also occur in environments with large predators, venomous animals, or high degrees of human conflict. In these instances, even one night away can involve substantial risk as is seen among the !Kung where a woman running away and spending the night outside of camp can be used to communicate “the strength of her feelings” about a an unhappy relationship [@shostak1981]\[\[6\]\]. When it comes to Utila, mothers in focus groups reported that running away on Utila was much less risky than it would be on mainland Honduras, with a few mothers mentioning that they warned their children about the difference in danger that came with running away on the mainland. Despite the relative safety of Utila, young children may still face risks out on their own due to proximity to the ocean, large dogs, and fast-moving motorcycles, golf carts, and tuk-tuks on the main road. For these reasons, running away may be too costly for young children but not costly enough to credibly signal need for older children.

*Sex and focal child alloparenting effort interact (postive for females, negative/flat for males)*

We found evidence of an interaction between how often a child cared for their siblings and the sex of the child in our models examining signal frequencies and cost, whereby there was a positive relationship between the frequency of and summed costliness of child signals of need and alloparenting effort for girls but no relationship or a slightly negative one for boys. In many societies, girls are often expected to do more alloparenting within the household and often end up doing so (cite). While girls in this sample were also reported to give more allocare, the difference was not significant. However, Utilian mothers reported in focus groups (confirm) that girls tend to provide more help for their siblings than boys. If this difference was the outcome of caretakers pressuring girls to do more (or resulted in the expectation of more alloparenting), the increase in signaling with alloparenting effort seen among girls may be the result of a strategy to signal the costliness of their efforts in order to invest more in romantic relationships and status seeking behaviors outside the household.

@@

@@

*More conflict with younger children but despite percieving older children as needy*

*Lack of other sex differences*

@@

There was a positive association between the summed frequency of child signaling and caretaker perceptions of child neediness within the family, something which would be consistent with sadness, crying, and temper tantrums being signals (or involving signals in the case of crying) if parents took this to mean how much they value additional investment.

@@

Parents were more likely to respond positively and less likely to respond negatively to the last time a child signaled when the cause involved discomfort, pain, injury, or illness than when it did not. One potential causal explanation for this is that parents may often have information for sources outside of child signaling to corroborate a child's need (e.g., a fever, a diagnosis, or witness a child hurt themselves) so that there is less private information about a child's current state. Simultaneously, conflicts of interest may be less significant in these times due to either 1) both sides benefiting similarly from their child returning to normal (e.g., with an illness) or 2) the cause involving something which is minor and would quickly resolve itself (e.g. a skinned knee). It may also be that many of the typical positive responses to pain and illness are not costly (e.g., providing comfort).

@@

One unexpected finding was a negative association between caretakers’ education level (years in school) and many of our child signaling measures despite our models controlling for household income and partner status. One possible reason for this may be that more educated caretakers see their children less during the day due to greater ability to afford childcare or due to them having jobs that are less compatible with child care (test model of social support predicted by education). Unfortunately, we do not have data on how much time caretakers spent with children to test this, and it seems equally likely that less educated parents may see their kids less often if they need to work more hours to support their families. Another possibility is that lower levels of education are associated with greater adversity for both parents and children, something which would increase the benefits of signaling need for children and the difficulty of parents to respond. It is also possible that education influences parenting norms in ways that cause differences in child signaling behavior.

@@

@@

Due to the fact the data used here were a part of survey which covered many topics and often multiple children, we sometimes faced a tradeoff between survey time and question quality. Asking adult participants about their relative investment in children in their household was likely not as effective as more in depth, single child focused questions due to both the relative lack of detail it provides and the potential for social desirability bias [@grimm2010], which could lead adult participants to be be more likely to report equal investment across children. Our questions about the last time the child signaled may have also been more effective if they focused on specific signals due to our methods combining crying, which is unambiguous as a signal, with sadness, which has both signaling and non-signaling functions.

Despite seeking data from both English-speaking participants born on the island Spanish-speaking immigrants from the mainland, our ability to detect differences between the populations may limited by our smaller sample of English-speaking Utilians. It is possible that differences in wealth and status between the groups translate to differences in child signaling due to children facing differerent types adversity or weighting different types of investment differently. It is also likely that the two groups differ in how caretakers could respond to child signaling, and caretakers between the two populations may also differ in their cultural models of parenting.

>>>>

## Positive assoication between child alloparenting effort and number of older children

@@

The effect of the number of older children in the household on the frequency of a child's alloparenting effort. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-alloparenting_dep_var for the corresponding regression table.
:::

@@

There were significant negative relationships between summed signal cost (`r stats$mBFc$BodyFat$str`) crying frequency (`r stats$mBFcry$BodyFat$str`), and tantrum frequency (`r stats$mBFtantrum$BodyFat$str`) and a scaled composite of subscapular and tricep skinfold thicknesses when controlling for the sex of the child, the age of the child, the frequency of conflict with one's primary caretaker, household demographics,caretaker perceptions of child immune function, an alloparenting effort by sex interaction and a body fat by sex interaction. See @tbl-bodyfat_models for regression statistics and @fig-effect_bodyfat for effects plots. Subscapular and tricep skinfold thicknesses were also significant predictors of summed signal cost and tantrum frequency when included in the same models as the sole measure of body fat. See xxxxSI TABLESxxxx for regression tables.

Although not significant, our models predicted a negative slope for the relationship between body fat and summed signal cost, crying frequency, and tantrum frequency for girls but no relationship for boys.

@@

We found no significant associations between sex and child signaling behavior outside of the previously mentioned interactions, including when sex was included in models as interaction with age. See table SI xxxxxxx for regression tables which included the sex by age interaction. Monthly household income also was not a significant predictor of signaling behavior in any model nor was population when using language, immigration status, and neighborhood as proxies for it. See table SI xxxxxxx.

@@

There was no significant association between the frequency of child sadness and age. Part of this likely results from the fact that, unlike crying, sadness is thought to have important non-signaling functions including drawing one’s attention to a problem, motivating deeper analysis of one’s situation, and causing withdrawal from situations with low fitness prospects (cite), from which, others can still obtain cue-based information about one's emotional state, well-being, and desires. By asking about sadness as a whole, our measure was unable to disentangle information caretakers gleaned from cues from information that came from signals. For this reason, a change in signals that arise from sadness outside of crying (e.g., facial sadness and some, but all, forms of withdrawal) may have been harder to detect with our sample size even if young children expressed facial sadness more often. It is also possible that there is no effect of age on sadness frequency among the age ranges examined it.

- add in depression over the lifecourse

@@

Consistent with signaling models which emphasize the role of conflicts of interest in favoring the use of costlier signals (cite this lab and others), there were positive associations between caretakers' perceptions of conflict frequency and 1) the summed costliness of children's signaling, 2) the frequency of child signals of need, and 3) the frequency of children's most common need signaling strategy. Although conflicts of interest do not always lead to behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest, making it a useful proxy for the underlying conflicts of interest that may increase the benefits of deception and the costs of being deceived\[\[5\]\]. If this is the case, the association between signaling and conflict may represent children's attempts at credibly signaling their value of additional investment or the costliness of punishment (cite) to parents who would have good reason to doubt the trustworthiness of simple verbal request (or the trustworthiness of facial sadness and crying for those resorting to temper tantrums).

Probably cut
When viewed in isolation, the frequencies of sadness, crying, and temper tantrums were also positively associated with conflict frequency between child and caretaker. However, the frequency of running away was not significantly associated with conflicts with parents. A retrospective study of parent-offspring conflict among Chon Chuuk immigrants in the northwest United States found that running away was the most common response to long-standing conflicts with parents and often resulted in beneficial outcomes for the child [@syme2023a]. The threat running away poses to child wellbeing and, therefore, its ability to credibly signal information about one’s current situation is likely highly dependent on one’s socioecology and age. Among the Chon Chuuk, running away is often seen as a suicide threat [@fischer1950; @syme2023a], likely due to xxxxxxxxx. Similar risks may also occur in environments with large predators, venomous animals, or high degrees of human conflict. In these instances, even one night away can involve substantial risk as is seen among the !Kung where a woman running away and spending the night outside of camp can be used to communicate “the strength of her feelings” about a an unhappy relationship [@shostak1981]\[\[6\]\]. When it comes to Utila, mothers in focus groups reported that running away on Utila was much less risky than it would be on mainland Honduras, with a few mothers mentioning that they warned their children about the difference in danger that came with running away on the mainland. Despite the relative safety of Utila, young children may still face risks out on their own due to proximity to the ocean, large dogs, and fast-moving motorcycles, golf carts, and tuk-tuks on the main road. For these reasons, running away may be too costly for young children but not costly enough to credibly signal need for older children.

*Sex and focal child alloparenting effort interact (positive for females, negative/flat for males)*

We found evidence of an interaction between how often a child cared for their siblings and the sex of the child in our models examining signal frequencies and cost, whereby there was a positive relationship between the frequency of and summed costliness of child signals of need and alloparenting effort for girls but no relationship or a slightly negative one for boys. In many societies, girls are often expected to do more alloparenting within the household and often end up doing so (cite). While girls in this sample were also reported to give more allocare, the difference was not significant. However, Utilian mothers reported in focus groups (confirm) that girls tend to provide more help for their siblings than boys. If this difference was the outcome of caretakers pressuring girls to do more or resulted in the expectation of more alloparenting, the increase in signaling with alloparenting effort seen among girls may be the result of a strategy to signal the costliness of their efforts in order to invest more in romantic relationships and status seeking behaviors outside the household.

@@

*Alloparenting effort model*
Not sure on interpretation here yet

@@

Children with less body fat cried more, threw more temper tantrums, and had higher summed signal costs. In other words, needier children signaled more often. Although the body fat by sex interaction was only marginally significant, our models predict a negative relationship between body fat and signaling for girls and no relationship for boys. Female reproduction is highly dependent on body fat, with conception requiring about xx% body fat and mothers investing a significant amount of calories in both children in the womb and breastfed infants (cite). For these reasons, low body fat is likely a be better predictor of future reproductive success in older girls than boys, potentially making signaling for increased investment a more viable strategy in instances when an individual has low body fat.

@@

There was a positive association between the summed frequency of child signaling and caretaker perceptions of child neediness within the family, something which would be consistent with sadness, crying, and temper tantrums being signals (or involving signals in the case of crying) if parents took this to mean how much they value additional investment. Caretakers also perceived older children as needier, despite having more conflicts with younger children. One reason for this could be that caretakers weighted monetary investment highly when evaluating child need.

@@

Parents were more likely to respond positively and less likely to respond negatively to the last time a child signaled when the cause involved discomfort, pain, injury, or illness than when it did not. One potential causal explanation for this is that parents may often have information for sources outside of child signaling to corroborate a child's need (e.g., a fever, a diagnosis, or witness a child hurt themselves) so that there is less private information about a child's current state. Simultaneously, conflicts of interest may be less significant in these times due to either 1) both sides benefiting similarly from their child returning to normal (e.g., with an illness) or 2) the cause involving something which is minor and would quickly resolve itself (e.g. a skinned knee). It may also be that many of the typical positive responses to pain and illness are not very costly (e.g., providing comfort).

@@

We found a negative association between caretakers’ education level (years in school) and many of our child signaling measures despite our models controlling for household income and partner status. One possible reason for this may be that more educated caretakers see their children less during the day due to greater ability to afford childcare or due to them having jobs that are less compatible with child care (test model of social support predicted by education). Unfortunately, we do not have data on how much time caretakers spent with children to test this, and it seems equally likely that less educated parents may see their kids less often if they need to work more hours to support their families. Another possibility is that lower levels of education are associated with greater levels of adversity for both parents and children, something which would increase the benefits of signaling need for children and the difficulty of parents to respond. It is also possible that education influences parenting norms in ways that cause differences in child signaling behavior.

@@

*Lack of evidence for demographic effects*

Outside of the sex by alloparenting effort interaction, we found little evidence of sex differences in child signaling behavior. As adults, females cry more than males, a finding that has been reported across cultures (cite). A similar sex difference is also seen with depression and suicide attempts, which some of us have argued to often be attempts at signaling need in other papers (cite), with females experiencing depression being about twice as often (cite) and young females being xxxx as likely to attempt suicide (normal citaiton). The most likely reason for this difference is that there is less of a sex difference in adversity, conflict, and cultural evaluations of signals of need children until adolescence and young adulthood. Although we do have data on adolescents, our small sample for this age range may have prevented us from detecting this.

Monthly income was also not found to be a significant predictor of signaling behavior. Many models designed to explain signals of need (e.g., the begging of young birds or withdrawal in humans) predict that offspring who are worse off or are receiving less investment have more to gain from signaling than those with more (cite). However, it is possible that this can be true and household level resources have little influence on child signaling frequencies between families. As signaling behavior is costly to child, caretakers, and siblings, children may gain direct and indirect fitness benefits from limiting their signaling behavior in times of resource stress (i.e., when the chances of gaining investment or the size of investment are small). Similarly, children in high resource families may also be motivated to signal at higher than expected frequencies due to the potential for larger pieces of investment. We know of no empirical explorations of this in humans (check more and confirm). However, multiple game theoretical models of sibling competition and have found that parental resources have a small impact on offspring signaling, especially with high rates of monogamy [Bossan + Mock and Parker + Check for newer sources]. Although counterintuitive, such an outcome fits the fact that fitness is relative and that natural selection should be expected to have favored those who maximized their own inclusive fitness, whether one found themselves in a great situation or a dire one.

Despite population level differences in wealth, job certainty, discrimination, and living conditions [@davis2014; @garcia2017; @garcia2018; @stonich2000], we generally did not find an effect of population on child signaling behavior. Although we sought data from both English-speaking participants born on the island and Spanish-speaking immigrants from the mainland, our smaller sample of English-speaking Utilians may have limited our ability to detect real differences between the populations. It is possible that differences in wealth and status between the groups translate to differences in child signaling due to children facing different types adversity or weighting different types of investment differently. It is also likely that the two groups differ in how caretakers could respond to child signaling, and caretakers between the two populations may also differ in their cultural models of parenting. However, as previously mentioned, it is also conceivable that household level differences have little effect on child signaling behavior due to the costs of signaling and a child's fitness interest in their relatedness.

@@

Due to the fact the data used here were a part of survey which covered many topics and often multiple children, we sometimes faced a tradeoff between survey time and question quality. Asking adult participants about their relative investment in children in their household was likely not as effective as more in depth, questions focused on specific questions. Framing the question this way may have also increased the likelihood of social desirability bias [@grimm2010], which could lead adult participants to be be more likely to report equal investment across children. Our questions about the last time the child signaled may have also been more effective if they focused on specific signals due to our methods combining crying, which is unambiguous as a signal, with sadness, which has both signaling and non-signaling functions.
